---
title: "Introduction to RDCOMClient"
author: "Jeff Lean"
date: '2018-10-22'
slug: introduction-to-rdcomclient
tags:
- RDCOMClient
- Excel
categories: []
---

For years I have been saying to myself - "I really should start a blog". Recently I stumbled across the [blogdown](https://bookdown.org/yihui/blogdown/) package and finally decided I had a simple enough tool to help me get started.

One Saturday night later and I am up and running with a functional blog. Fast forward a few months and here we are. I've decided that the best topic for me to write about first is the [RDCOMClient](http://www.omegahat.net/RDCOMClient/Docs/introduction.html) package.

I've been using this package for the last few years in my work. When I was first exposed to it, I'll admit I was a bit reluctant to use it because I didn't understand the syntax and I couldn't find many examples online. It took a while to get used to using it, and I can't say it was a simple process, but now that I have used it for a while it is one of my favourite packages. It allows you to drive almost every aspect of Excel from within an R script. This makes constructing very detailed outputs possible with complete reproducibility.

RDCOMClient is written by Duncan Temple Lang. You can find an introduction to the package here - [RDCOMClient](http://www.omegahat.net/RDCOMClient/Docs/introduction.html). 

I started off using RDCOMClient to read some csv documents that had unusual formatting. This rapidly progressed into reading groups of xls documents with variable formatting. I needed to use the Excel Find() function to know what row and column to start on. Being able to target the precise range based on values in the data was very helpful. Eventually I built a script that took a handful of seed documents as inputs and crawled across a network of linked Excel documents to produce a picture of over 500 linked spreadsheets on a network.

This post will be the first in a series of posts about how to use RDCOMClient. I hope to build a repository of examples that anyone can use to learn how to use this brilliant package.

The very first thing to do is to install the package.

```{r, eval=FALSE}
install.packages("RDCOMClient", repos = "http://www.omegahat.net/R")
```

You'll also need a working version of Microsoft Excel, so I'll just assume that you do. 

First you need to open a connection to the COM object

```{r, eval=TRUE}
library(RDCOMClient)
ex <- COMCreate("Excel.Application")
```

So far it doesn't look like much is happening does it? Well we've only just created an instance of Excel. We haven't actually opened the Excel interface. Excel is actually running now, but we just can't see it.


```{r, eval=TRUE}
ex[["Visible"]]
```












