---
title: RDCOMClient Part 5 - Extending colour formatting
author: Jeff Lean
date: '2018-10-31'
slug: rdcomclient-part-5-extending-colour-formatting
categories:
  - Excel
  - R
  - RDCOMClient
tags:
  - Excel
  - R
  - RDCOMClient
---

```{r, eval=TRUE, echo=FALSE, messages = FALSE}
library(RDCOMClient)
suppressMessages(library(dplyr))
library(purrr)

ex <- COMCreate("Excel.Application")
ex[["Visible"]] <- TRUE

Workbook <- ex[["workbooks"]]

current_file <- Workbook$Add()

sheet <- current_file$Worksheets(1)

range <- sheet$Range("A1")

myString <- "Hello World!"

range[["Value"]] <- myString

dataCellFont <- range[["Font"]]

dataCellFont[["Size"]] <- 20

dataCellFont[["Bold"]] <- TRUE

```

In my last post I introduced creating a new Excel workbook, writing data into that workbook, and applying some formatting to that data. In this post I'll cover the following:

* Simplifying the colour specification
* Saving your workbook

## Simplifying colour specification

As we saw in the previous post, it is a bit difficult telling Excel what colour to use for something. You need to know the exact numeric reference for a colour and unless you have identified the colour through trial and error previously, it's just not something you want to be doing.

Luckily for us this problem has been largely solved already. We will need to introduce another package here called [gplots](https://www.rdocumentation.org/packages/gplots/versions/3.0.1). This package includes a function called `col2hex`. This function takes a colour string like "red", and outputs the respective hexadecimal code. 

According to [this link](https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/colorPaletteCheatsheet.pdf), R has 657 built-in colour names. We can just pass the desired colour name to the `col2hex` function and get our hex code back.

About now some of you are probably asking what good is a hexadecimal code since we can't pass it directly to Excel. Excellent question, well presented! When I first started playing around with this problem, I found the `strtoi` function from base R that converts strings to integers with a specified base. I thought I could just pass the hex code to this function and I'd easily get a number back. Well as it happens, you do get a number back - it's just not the right number if you pass the whole hex code to the function.

Here's what I tried first. I'll assume that you've got your Excel document open using the method we used in the previous post.

```{r, eval=TRUE, message = FALSE, warning=FALSE}
suppressMessages(library(gplots))
library(stringr)

# try converting a known colour to a hexadecimal value
col2hex('red') # should give "#FF0000"

# try converting that hexadecimal string to a decimal number
strtoi("FF0000") 

# that didn't work because we didn't specify the base
strtoi("FF0000", base = 16) 

```

Since I knew that the colour red is represented by the hex code "#FF0000" I expected to see the strtoi() function return a value of 255, since I also knew that this is the number that red should correspond to. Clearly 16711680 was not going to give me the result I wanted. After a few hours looking around online I found [this post](https://stackoverflow.com/questions/8468855/convert-a-rgb-colour-value-to-decimal) that described how the hex codes represent the red, green and blue channels of a colour.

After finding that little gem, I just wrote a simple little function that takes a colour name as the argument. This function converts that string to a hexadecimal code, converts the respective colour channels to decimal numbers, and then combines them in the necessary way to arrive at a decimal number we can pass to Excel through the `[["Color"]]` property.

```{r, eval=TRUE, messages = FALSE}

# create our function that will be used to calculate the decimal
# representation of a colour
colNum <- function(colour = "black"){
  
  # catch error if a value other than a recognised colour is entered
  if(!(colour %in% colours())){
    stop("Please enter one of the R specified colours as a string.\n
         The R colours can be found by typing colours().")
  }
  
  # first convert colour to hexadecimal code and strip out the hashtag
  value <- col2hex(colour) %>%
    str_replace("#", "")
  
  # get the component colours
  R <- str_sub(value, 1, 2) %>% strtoi(base = 16)
  
  G <- str_sub(value, 3, 4) %>% strtoi(base = 16)
  
  B <- str_sub(value, 5, 6) %>% strtoi(base = 16)
  
  # calculate the decimal resultant of the three colour channels
  colour <- B * 65536 + G * 256 + R
  
  return(colour)
}

# now we can specify the colour of text in our target cell simply by 
# passing a colour name to our function
dataCellFont[["Color"]] <- colNum("magenta1")

dataCellFont[["Color"]] <- colNum("springgreen2")

dataCellFont[["Color"]] <- colNum("red")

```


## Saving your workbook

Now that you've gone through all the effort to create your "Hello World!" document, you'll be wanting to save it so you can keep it.

This is a simple matter of using the SaveAs method on the workbook object. Be sure to specify where you want to save your document by passing a path string as an argument. If you don't provide a path, the document will save in your current working directory. In order to make sure your document saves smoothly, you can include the command `ex[["DisplayAlerts"]] <- FALSE` before you save your document. This will prevent a popup message asking if you want to replace your document. It is important to remember to restore your DisplayAlerts property to TRUE at the end of your script.


```{r, eval=TRUE, messages = FALSE}
ex[["DisplayAlerts"]] <- FALSE
current_file$SaveAs("D:\\notpeerreviewed\\data\\hello_world.xlsx")

# close the current document
current_file$Close(SaveChanges = TRUE)
ex[["DisplayAlerts"]] <- TRUE

# now we can quit Excel
ex$Quit()
```

This post has been a bit more about understanding the underlying workings of how R deals with colours and how different colour codes work. In my next post I'll show how to write a more realistic example of data into a workbook.

Session info for this post.
```{r, eval=TRUE, message=FALSE}
print(sessionInfo(), locale = FALSE)

```







